---
title: "ARIMA BAseline for forecasting S&P 500 index"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    number_sections: false
    toc: yes
    toc_depth: 3
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '3'
---

```{r init, include=FALSE}
# Load all required packagess
library(xts)
library(ezids)
library(dplyr)
library(tseries)
library(ggplot2)
library(forecast)
library(corrplot)
library(lubridate)

knitr::opts_chunk$set(warning = F, results = "markup", message = F)
options(scientific=T, digits = 3) 
```


# Data Loading and Preprocessing.
```{r, include=F}
df <- data.frame(read.csv('https://raw.githubusercontent.com/DATS6101-TeamNeo/datasets/main/dataset2003-2023.csv'))
df$Date <- as.Date(df$Date)
str(df)
work_df <- subset(df, select = -c(6))
str(work_df)
summary(work_df)
```

```{r, include=T, results=T}
work_df$daily_change <- (work_df$Close / lag(work_df$Close) - 1) * 100
head(work_df)
work_df <- na.omit(work_df)
head(work_df)
```

### Check if the time series is Stationary.
```{r, include=T, results=T}

adf.test(work_df$Close)

# Check the stationarity using the ACF and PACF plots.
acf(work_df$Close, main = "ACF Plot")
pacf(work_df$Close, main = "PACF Plot")
```

### Making the time series data stationary
Using first order differencing to make the time series data stationary.
```{r, include=T, results=T}

ts_data <- xts(work_df$Close, order.by = work_df$Date)
```
ADF test for the original time series data.
```{r, include=T, results=T}
adf.test(ts_data)

diff_ts_data <- diff(ts_data, differences = 1)

diff_ts_data <- na.omit(diff_ts_data)
```
ADF test for the differenced time series data.
```{r, include=T, results=T}
adf.test(diff_ts_data)
```

```{r, include=T, results=T}
acf(diff_ts_data, main = "ACF Plot")
pacf(diff_ts_data, main = "PACF Plot")
```

# ARIMA model fitting and forecasting
From the adf test and ACF and PACF plots, we can see that the time series data is close to stationarity. We can now create a baseline ARIMA model and forecast the future values.
```{r, include=T, results=T}

train_data <- head(diff_ts_data, n = round(0.98 * length(diff_ts_data)))
test_data <- diff_ts_data[-seq_along(train_data)]


arima_model <- auto.arima(train_data)


summary(arima_model)


forecasts <- forecast(arima_model, h = length(test_data))
```

# Evaluation of the ARIMA model
```{r, include=T, results=T}

# Convert the forecasts to the original scale.
last_train_date <- index(tail(train_data, 1))
last_observed_value <- ts_data[last_train_date]
actual_preds <- as.vector(last_observed_value) + cumsum(forecasts$mean)
actual_test <- as.vector(last_observed_value) + cumsum(test_data)
actual_preds <- xts(actual_preds, order.by = index(actual_test))

# Compute the forecasting errors
errors <- actual_preds - actual_test

# Calculate evaluation metrics
mae <- mean(abs(errors))
mse <- mean(errors^2)
rmse <- sqrt(mse)
```

Mean Absolute Error (MAE): `r mae`

Mean Squared Error (MSE): `r mse`

Mean Squared Error (RMSE): `r rmse`

# Plots
### Plot train and test data.
```{r, include=T, results=T}
plot(ts_data[index(ts_data) <= last_train_date], type = "l", xlab = "Time", ylab = "Index closing price", main = "Train Data")
plot(ts_data[index(ts_data) > last_train_date], type = "l", xlab = "Time", ylab = "Index closing price", main = "Test Data")
```

### Plot the forecasted data along with the train and test data.

```{r, include=T, results=T}

plot_start_date <- as.Date("2023-01-01")

actual_train <- ts_data[index(ts_data) > plot_start_date & index(ts_data) <= last_train_date]

# Convert the time series data to data frames
train_df <- data.frame(Date = index(actual_train), Value = coredata(actual_train), Type = "Train data")
test_df <- data.frame(Date = index(actual_test), Value = coredata(actual_test), Type = "Test data")
forecast_df <- data.frame(Date = index(actual_preds)[-1], Value = actual_preds[-1], Type = "Forecast")

# Combine the data frames
combined_df <- rbind(train_df, test_df, forecast_df)

# Create the plot
ggplot(combined_df, aes(x = Date, y = Value, color = Type, linetype = Type)) +
  geom_line() +
  labs(title = "Train, Test, and Forecast Data Plot",
       x = "Date",
       y = "Closing Price") +
  scale_color_manual(values = c("red", "orange", "blue")) +
  scale_linetype_manual(values = c("dashed", "solid", "solid")) +
  theme_minimal() +
  theme(legend.position = "right")
```